---
title: "Data Management and Visualisation: Forest Research"
output:
  html_document:
    df_print: paged
    fig_width: 7
    fig_height: 6.37
    fig_caption: true
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: show
  pdf_document:
    df_print: default
    fig_width: 7
    fig_height: 6.37
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## The Growth of Woodland Area within the United Kingdom (1998-2024)

### Data Source

This project is focused on visualising various changes in woodland within the United Kingdom between March 31st of 1998 and 2024 respectively. The raw data and logo for this project were retrieved from ['Forest Research'](https://www.forestresearch.gov.uk/tools-and-resources/statistics/data-downloads/), a branch of the Forestry Commission (FC) which collects data by totaling extractions made from multiple forest services operating within the United Kingdom. It principally gathers information on tree-related statistics, including but not limited to gathering longitudinal provisional data on the estimated amount of woodland area in the UK.

The “Woodland area, UK, 1998 to 2024” ODS file contains 5 sheets of actual data, though was supplemented with cover, content and notes pages. In tandem, the pages containing data were annotated with further notes clarifying the origin of the data and different commissions involved in gathering it. The pages containing data split woodland area based on whether it was coniferous or deciduous, private or public sector, and reported total estimated woodland per country and in the United Kingdom as a whole.

```{r, message=FALSE, warning=FALSE, echo=FALSE}

## Should the below list of packages need to be installed, remove '#'(s) and run:
#libraries<-c("tidyverse", "cowplot", "magick", "readODS", "here", "plotly", "highcharter", "htmlwidgets")
#install.packages(libraries, repos="http://cran.rstudio.com")

## NECESSARY PACKAGES:

library(here)         # to create user-specific filepaths
library(readODS)      # to retrieve data from ODS files
library(tidyverse)    # for various visualisation packages
library(cowplot)      # for various visualisation options
library(magick)       # to import images onto visualisations
library(plotly)       # to create interactive scatterplots
library(highcharter)  # to create interactive barplots
library(htmlwidgets)  # to save interactive visualisations

## DATA PREPARATION:
# Extraction of raw data from ODS file

country_names<-c("England","Wales","Scotland","Northern Ireland")
pathway<-paste0(here("raw_secondary_data", "area-timeseries-20jun24.ods"))
countries_list<-list()
for (i in seq_along(country_names)){
  countries<-country_names[i]
  countries_list[[countries]]<-read_ods(pathway, sheet = i+3)
} # this loop extracts all of the raw data from the country-specific sheets and imports it into R
raw_extracted_data<-data.frame(countries_list) # which is then put into a dataframe
```

### Glimpse of the Raw Data ...

```{r, message=FALSE, warning=FALSE, echo=FALSE}
# Glimpse of the raw data

head(raw_extracted_data)
```

### Now after Processing ...

```{r, message=FALSE, warning=FALSE, echo=FALSE}
# Processing and cleaning of data

processed_extracted_data<-raw_extracted_data[-c(1:3),] # removes unnecessary text
names(processed_extracted_data)<-as.matrix(processed_extracted_data[1,]) # labels each column by their original titles
processed_extracted_data<-processed_extracted_data[-1,] # removes text names from the data
names(processed_extracted_data)<-make.unique(names(processed_extracted_data)) # makes each column unique
processed_extracted_data<-processed_extracted_data %>% select(-starts_with("Note")) # removes erroneous "note" columns
processed_extracted_data<-processed_extracted_data%>% mutate_if(is.character, as.numeric) # converts all variables into numeric form
rownames(processed_extracted_data) <- NULL # corrects for the erroneous row numbers extracted when subsetting from the raw data

# Glimpse of the processed data
head(processed_extracted_data)

```

### The Research Question(s)

The main rationale for exploring this data was to elucidate various visual patterns within the growth of woodland area across the United Kingdom from 1998 to 2024. This includes creating a static visualisation of absolute numeric increase, annotated with the rate (*percentage increase*) of development, as well as more advanced interactive visualisations which show specific patterns in *coniferous* or *deciduous*, *public* or *private* woodland.

Calculating the *percentage increase* was important to contextualise the proportional rate of growth for each country, beyond simply interpreting development based on raw numeric increase in hectares of woodland. Furthermore, following this up with interactive visualisations allowed more precise dissemination of whether patterns of development were distinct depending on *tree type* (i.e. *coniferous* or *deciduous*) and woodland ownership (i.e. *public* or *private land*).

```{r, message=FALSE, warning=FALSE, echo=FALSE}
# Mutual code for all visualisations
fig_path<-here("figs") # creates the necessary path for saving the figures
logo_file<-paste0(here("logo","Picture1.jpg")) # creates the path for applying the logo
year_ending_March_31st<-processed_extracted_data$`Year ending 31 March` # timeframe for visualisations

# The first plot, aptly named 'Basic Plot', was created while at the University of Sheffield as part of initial novice attempts at writing code, examined during the PSY6422 Module.

## DATA EXTRACTION: 
# Extraction of necessary variables from processed dataframe

country_names<-c("England", "Wales", "Scotland", "Northern Ireland")
woodland_area<-select(processed_extracted_data, ends_with(" total (thousand ha)")) # Total Woodland Area
for (country in country_names){
  assign(country, woodland_area[[paste0(country, " total (thousand ha)")]])
} # creates numeric variables for the woodland area of each country, to be labelled per year

woodland_area<-data.frame(England,Wales,Scotland,`Northern Ireland`) # which is then put back into the dataframe
woodland_area<-woodland_area%>% rename(Northern_Ireland=`Northern.Ireland`) # *corrects for interaction issues now caused by spacing in `Northern Ireland`
country_names<-c("England", "Wales", "Scotland", "Northern_Ireland") # *

woodland_by_year<-cbind(year_ending_March_31st,woodland_area) # Total Woodland Area, By Year

# Calculation of proportional change in woodland area per country
percentage_results<-data.frame(country=character(),percentage_increase=numeric(),stringsAsFactors=FALSE) # creates an empty dataframe
for(country in country_names){
  min_v<-min(woodland_area[[country]])
  max_v<-max(woodland_area[[country]])
  percentage_increase<-(((max_v-min_v)/min_v)*100)
  percentage_results<-rbind(percentage_results, data.frame(country=country,percentage_increase=percentage_increase))
} # this loop calculates the percentage increase in woodland area of each country from 1998-2024
percentage_results[1:4,2]<-round(percentage_results[1:4,2],2) # then rounds the data to 2 decimal places

## FINAL DATAFRAME:
woodland_growth_over_time<-data.frame(
  country=c(rep("England",27),rep("Wales",27),rep("Scotland",27),rep("Northern Ireland",27)),
  woodland=c(woodland_by_year$England,woodland_by_year$Wales,
             woodland_by_year$Scotland,woodland_by_year$Northern_Ireland),
  year=c(woodland_by_year$year_ending_March_31st)) # creates a final dataframe amenable to the upcoming visualisation

mapping<-aes(x=year,y=woodland,colour=country) # creates the mapping for the visualisations

## BASIC STATIC PLOT:
# Creates a plot mapping the amount of woodland area development from 1998-2024, as divisable by country, and as annotated with percentage increase

BasicPlot<-woodland_growth_over_time %>%
  ggplot(mapping=mapping)+
  geom_smooth(method="gam")+
  labs(x="Year (commencing from March 31st)",
       y="Woodland area (in thousand hectares)",
       colour="Country",
       title="The Growth of Total Woodland Area within the United Kingdom",
       subtitle="As annotated with percentage increase from 1998-2024",
       caption="Data retrieved from: Forest Research, 2024")+
  annotate("label",x=2010,y=1210,label=paste(percentage_results[1,2],"%"),colour="#EE0000",size=3,fontface="bold")+
  annotate("label",x=2010,y=375,label=paste(percentage_results[2,2],"%"),colour="#00CD00",size=3,fontface="bold")+
  annotate("label",x=2010,y=1450,label=paste(percentage_results[3,2],"%"),colour="#0000CD",size=3,fontface="bold")+
  annotate("label",x=2010,y=175,label=paste(percentage_results[4,2],"%"),colour="#FFA500",size=3,fontface="bold")+
  scale_colour_manual(values=c(England="#EE0000",Wales="#00CD00",
                               Scotland="#0000CD",`Northern Ireland`="#FFA500"))+
  scale_x_continuous(breaks=seq(1998,2024,5))+
  scale_y_continuous(breaks=seq(0,1500,150))+
  theme(panel.border=element_rect(colour="#8B7355",fill=NA,linewidth=2),
        panel.grid.minor=element_line(colour="#8B7355",linewidth=0.5),
        panel.grid.major=element_line(colour="#CAFF70",linewidth=0.7),
        panel.background=element_rect(fill="#FFFFF0"),
        axis.line=element_line(linewidth=2,colour="#8B7355"),
        plot.title=element_text(face="bold"),
        plot.subtitle=element_text(face="italic"),
        text=element_text(family="serif"),
        legend.title=element_text(face="bold"),
        legend.box.background=element_rect(colour="#8B7355"),
        legend.box.margin=margin(1,1,1,1),
        legend.key=element_rect(colour="#8B7355"))

# To demonstrate the ability to do so, this code adds the official logo to sit alongside the data source

BasicPlot<-ggdraw(BasicPlot)+
  draw_image(logo_file, scale=.2,x=1,hjust=1,halign=1,valign=0)

## Visualisation of the Growth of Total Woodland Area within the United Kingdom, from 1998 to 2024

BasicPlot

```

While the previous basic plot was initially taken from a Data Science Project completed during my Master's studies, the following code demonstrates a more recent attempt at configuring and advancing upon my previous abilities with data management and visualisation. Also, now that we've already gone through the process of showing a more general composition of woodland development from 1998-2024, it could be curious to investigate how this could differ based on *land ownership* and *tree type*.

```{r, message=FALSE, warning=FALSE, echo=FALSE}
# DATA EXTRACTION: Woodland Tree Types on PUBLIC land ...

public_trees<-data.frame(
  tree_type=c(rep("FE conifers, England",27),rep("FE broadleaves, England",27),rep("NRW conifers, Wales",27),rep("NRW broadleaves, Wales",27),
              rep("FLS conifers, Scotland",27),rep("FLS broadleaves, Scotland",27),rep("FS conifers, NI",27),rep("FS broadleaves, NI",27)),
  woodland=c(processed_extracted_data$`FE conifers (thousand ha)`,processed_extracted_data$`FE broadleaves (thousand ha)`,
             processed_extracted_data$`NRW conifers (thousand ha)`,processed_extracted_data$`NRW broadleaves (thousand ha)`,
             processed_extracted_data$`FLS conifers (thousand ha)`,processed_extracted_data$`FLS broadleaves (thousand ha)`,
             processed_extracted_data$`FS conifers (thousand ha)`,processed_extracted_data$`FS broadleaves (thousand ha)`),
  year=c(`year_ending_March_31st`)) # creates a dataframe amenable to the upcoming visualisation

public_trees$tree_type<-factor(public_trees$tree_type, levels=c("FE conifers, England", "FE broadleaves, England", 
                                                                  "NRW conifers, Wales", "NRW broadleaves, Wales",
                                                                  "FLS conifers, Scotland", "FLS broadleaves, Scotland",
                                                                  "FS conifers, NI", "FS broadleaves, NI")) # this code facilitates consistent ordering of variables in visualisation legends

mapping<-aes(x=year,y=woodland,colour=tree_type) # creates the mapping for the visualisations

## BASIC INTERACTIVE PLOT 1:

# Generates an interactive scattergraph of documented PUBLIC coniferious and deciduous woodland area within the United Kingdom from 1998-2024


InteractivePlot_public<-public_trees %>%
  ggplot(mapping=mapping)+
  geom_point(aes(text = paste("Woodland Area:", woodland, "(K ha)",
                               "<br>Year:", year, "<br>Type & Region:", tree_type), shape=tree_type))+
  labs(title="The Growth of Coniferous and Deciduous (Broadleaf) Public Woodland Area in the United Kingdom",
       x="Year (commencing from March 31st)",
       y="Woodland area (in thousand hectares)",
       colour="Woodland Type and Region",
       shape=NULL)+
  scale_colour_manual(values=c(`FE conifers, England`="#FF0000",`FE broadleaves, England`="#FF0000",
                               `NRW conifers, Wales`="#00CD00",`NRW broadleaves, Wales`="#00CD00",
                               `FLS conifers, Scotland`="#0000CD",`FLS broadleaves, Scotland`="#0000CD",
                               `FS conifers, NI`="#FFA900",`FS broadleaves, NI`="#FFA900"))+
  scale_shape_manual(values=c(0,15,1,16,2,17,5,18))+
  scale_x_continuous(breaks=seq(1998,2024,4))+
  scale_y_continuous(breaks=seq(0,1000,50))+
  theme(panel.border=element_rect(colour="#8B7355",fill=NA,linewidth=2),
        panel.grid.major=element_line(colour="beige",linewidth=0.7),
        axis.line=element_line(linewidth=2,colour="#8B7355"),
        panel.background=element_rect(fill="beige"),
        plot.title=element_text(face="italic"),
        text=element_text(family="sans",size=12))

# Unfortunately, the logo cannot be added to an interactive plot as this feature has not yet been implemented into plotly

# As proper preparations and modifications have already been made, the plot can now instantly be made interactive

InteractivePlot_public<-ggplotly(InteractivePlot_public, tooltip="text",width=900,height=600) %>%
  layout(margin = list(t = 50, r = 50, b = 50, l = 50))

## Visualisation of the Growth of Coniferous and Deciduous Public Woodland Area within the United Kingdom, from 1998 to 2024

InteractivePlot_public
```

```{r, message=FALSE, warning=FALSE, echo=FALSE}
# DATA EXTRACTION: Woodland Tree Types on PRIVATE land ...

private_trees<-data.frame(
  tree_type=c(rep("Private sector conifers, England",27),rep("Private sector broadleaves, England",27),rep("Private sector conifers, Wales",27),rep("Private sector broadleaves, Wales",27),
              rep("Private sector conifers, Scotland",27),rep("Private sector broadleaves, Scotland",27),rep("Private sector conifers, NI",27),rep("Private sector broadleaves, NI",27)),
  woodland=c(processed_extracted_data$`Private sector conifers (thousand ha)`,processed_extracted_data$`Private sector broadleaves (thousand ha)`,
             processed_extracted_data$`Private sector conifers (thousand ha).1`,processed_extracted_data$`Private sector broadleaves (thousand ha).1`,
             processed_extracted_data$`Private sector conifers (thousand ha).2`,processed_extracted_data$`Private sector broadleaves (thousand ha).2`,
             processed_extracted_data$`Private sector conifers (thousand ha).3`,processed_extracted_data$`Private sector broadleaves (thousand ha).3`),
  year=c(`year_ending_March_31st`)) # creates a dataframe amenable to the upcoming visualisation

private_trees$tree_type<-factor(private_trees$tree_type, levels=c("Private sector conifers, England", "Private sector broadleaves, England", 
                                                                  "Private sector conifers, Wales", "Private sector broadleaves, Wales",
                                                                  "Private sector conifers, Scotland", "Private sector broadleaves, Scotland",
                                                                  "Private sector conifers, NI", "Private sector broadleaves, NI")) # facilitates consistent legend order

## BASIC INTERACTIVE PLOT 2:

# Generates an interactive scattergraph of documented PRIVATE coniferious and deciduous woodland area within the United Kingdom from 1998-2024

InteractivePlot_private<-private_trees %>%
  ggplot(mapping=mapping)+
  geom_point(aes(text = paste("Woodland Area:", woodland, "(K ha)",
                              "<br>Year:", year, "<br>Type & Region:", tree_type), shape=tree_type))+
  labs(title="The Growth of Coniferous and Deciduous (Broadleaf) Private Woodland Area in the United Kingdom",
       x="Year (commencing from March 31st)",
       y="Woodland area (in thousand hectares)",
       colour="Woodland Type and Region",
       shape=NULL)+
  scale_colour_manual(values=c(`Private sector conifers, England`="#FF0000",`Private sector broadleaves, England`="#FF0000",
                               `Private sector conifers, Wales`="#00CD00",`Private sector broadleaves, Wales`="#00CD00",
                               `Private sector conifers, Scotland`="#0000CD",`Private sector broadleaves, Scotland`="#0000CD",
                               `Private sector conifers, NI`="#FFA900",`Private sector broadleaves, NI`="#FFA900"))+
  scale_shape_manual(values=c(0,15,1,16,2,17,5,18))+
  scale_x_continuous(breaks=seq(1998,2024,4))+
  scale_y_continuous(breaks=seq(0,1000,100))+
  theme(panel.border=element_rect(colour="#8B7355",fill=NA,linewidth=2),
        panel.grid.major=element_line(colour="beige",linewidth=0.7),
        axis.line=element_line(linewidth=2,colour="#8B7355"),
        panel.background=element_rect(fill="beige"),
        plot.title=element_text(face="italic"),
        text=element_text(family="sans",size=12))

InteractivePlot_private<-ggplotly(InteractivePlot_private, tooltip="text",width=900,height=600) %>%
  layout(margin = list(t = 50, r = 50, b = 50, l = 50))

## Visualisation of the Growth of Coniferous and Deciduous Private Woodland Area within the United Kingdom, from 1998 to 2024

InteractivePlot_private
```

Generally, these visualisations tend to indicate that private woodland area has grown at a much more visually noticeable rate than public woodland, even after accounting for the differences in initial proportions by creating graphs with relative scales. This could be curious for motivating investigations into WHY this is happening (i.e. privatisation of forested areas for production purposes, higher funding for development, etc.) or even whether this difference is statistically significantly different in the first place. Regardless, it was entertaining to play around with the data to show this.

And just as a final tidbit, there is an interactive barplot below showing all of the used data, combining private and public woodland statistics to show a more general trend on the total amount of woodland area as denoted by whether it was coniferous or deciduous, per country, per year from 1998-2024. This could also be interpreted as a more advanced version of our original static plot. :) 

```{r, message=FALSE, warning=FALSE, echo=FALSE}

# DATA MANIPULATION: Woodland Tree Types on both PUBLIC and PRIVATE land ...

total_trees<-cbind(public_trees,private_trees) # combines all data into one dataframe
names(total_trees)<-make.names(names(total_trees), unique=TRUE) # renders names unique to allow renaming
total_trees<-total_trees%>% 
  rename(private_woodland=`woodland.1`,public_woodland=woodland)%>% # renames variables to facilitate interaction
  mutate(sources=paste(tree_type,`tree_type.1`, sep=" + "))%>% # creates variable to record source of woodland area
  select(-c(1,4,6))%>% # removes now unneccessary variables from the dataframe
  select(c("sources", "year", "public_woodland", "private_woodland")) # reorders dataframe to improve readability
total_trees$total_woodland<-rowSums(total_trees[, c("public_woodland","private_woodland")], na.rm=TRUE) # calculates total woodland areas

visual_theme<-hc_theme( # creates an aesthetically similar "woodland"-esque theme to the prior visualisations
  colors=c("#FF0000","darkred","#00CD00","darkgreen","#0000CD","darkblue","#FFA900","darkorange"),
  chart=list(plotBackgroundColor="beige",
             style=list(fontFamily="Helvetica Neue", fontSize="15px")),
  title=list(style=list(fontWeight="bold",fontSize="20px")),
  yAxis=list(gridLineColor="#8B7355"))

# BASIC INTERACTIVE PLOT 3:

# Generates an interactive stacked barplot using the 'highcharter' rather than 'plotly' package, showing totals of coniferious and deciduous woodland area within the United Kingdom from 1998-2024

amalgamated_visualisation<-total_trees %>% 
  hchart('column', hcaes(x=year,y=total_woodland,group=sources),stacking="normal")%>%
  hc_title(text="Total Woodland Area of the United Kingdom from 1998-2024")%>%
  hc_subtitle(text="Segmented by tree type (coniferous or deciduous) and country")%>%
  hc_xAxis(title=list(text="Year (commencing from March 31st)"))%>%
  hc_yAxis(title=list(text="Woodland area (in thousand hectares)"))%>%
  hc_add_theme(visual_theme)%>%
  hc_caption(text="Northern Ireland only began documenting woodland area as of 2005: see abline")%>%
  hc_annotations(list(shapes=list(list(type = 'path',points = list(list(xAxis = 0, yAxis = 0, x = 2004.5, y = 0),
                                                                   list(xAxis = 0, yAxis = 0, x = 2004.5, y = 3500)),
                                       stroke="black",strokeWidth=1)),draggable=""))

# Trend in RECORDED woodland area (i.e. as 1998-2004 NI woodland area was not documented) in the United Kingdom from 1998-2024.

amalgamated_visualisation
```

### Full Data Reference

Forest Research (2024, March 31). *Tools and Resources: Data Downloads*. https://www.forestresearch.gov.uk/tools-and-resources/statistics/data-downloads/
